Intro Statistical Analysis Script

### 2024-08-15 code/analysis-code/fullanalysis-v1.qmd
Revisions for submission 2
After call w/Bill

whereas BW is birth weight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.

@@

# Create a gtsummary table directly from the model
table_model1 <-
  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model1

@@

whereas BW is birthweight in grams, MA is maternal ancestry (Euro-descent, Afro-descent), Gest is gestational status (nulliparous or multiparous), Age is maternal age in years, and $\epsilon$ is the error term. Results are displayed in Table 3.

```{r}
# Final model 2 for paper using all data
# Using recipes and workflows
# Define recipe for model
ML_recipe2 <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
  #step_dummy(ModifiedColor) #%>% # Convert all nominal variables to dummy variables. Cannot do this if I am merging tables in gtsummary
  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors. Cannot do this if I am merging tables in gtsummary

@@

# Create a gtsummary table directly from the model
table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model2

@@

```{r}
# Combine the two tables using tbl_merge
table3 <- tbl_merge(list(table_model1, table_model2), tab_spanner = c("**Model 1**", "**Model 2**")) %>%
  modify_table_body(fun = ~.x %>% arrange(variable))
table3

@@

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$
whereas BW is birthweight in grams, MC is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

```{r}
# Final model 3 for paper using all data
# Using recipes and workflows
# Simple linear model
# Specify recipe
ML_recipe3 <- recipe(Weightgrams ~ Color, data = ML_linear) %>%
  step_relevel(Color, ref_level = "White") #%>% # Set reference level for White
  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors

@@

# Create a gtsummary table from model
table_model3 <- tbl_regression(ML_fit3, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model3

@@

whereas BW is birthweight in grams, MC is maternal skin color (White, Black and mixed race), Gest is gestational status (nullipara, primipara, multipara, multigravida), Nat is maternal nationality (Brazilian, Latin American, European, Middle Eastern), Age is maternal age in years, and $\epsilon$ is the error term.

```{r}
# Final model 4 for paper using all data
# Using recipes and workflows
# Complex linear model
# Specify recipe
ML_recipe4 <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
  step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
  step_relevel(ModifiedStatus, ref_level = "Multiparous") #%>% # Set reference level for ModifiedStatus
  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors

@@

# Create a gtsummary table from model
table_model4 <- tbl_regression(ML_fit4, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>% # Hide confidence intervals and p-values
    add_glance_table(include = c(r.squared, nobs)) # Add R-squared and number of observations
table_model4

>>>>

whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.

@@

# Create a gtsummary table directly from the model
table_model1 <-
  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs)) %>%
  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
table_model1

@@

whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term. Results are displayed in Table 3.

```{r}
# Model 2: Final model for paper using all data
# Using recipes and workflows
# Define recipe for model
ML_recipe2 <- recipe(Weightgrams ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") #%>% # Set reference level for ModifiedColor
  #step_dummy(ModifiedColor) #%>% # Convert all nominal variables to dummy variables. Cannot do this if I am merging tables in gtsummary
  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors. Cannot do this if I am merging tables in gtsummary

@@

# Create a gtsummary table directly from the model
table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs)) %>%
  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
table_model2

@@

```{r}
# Combine the two tables using tbl_merge
table3 <- tbl_merge(
  list(table_model1, table_model2),
  tab_spanner = c("**Model 1**", "**Model 2**")) %>%
  modify_table_body(fun = ~.x %>% arrange(variable))
table3

@@

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$
whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

```{r}
# Model 3: Final model for paper using all data
# Using recipes and workflows
# Simple linear model
# Specify recipe
ML_recipe3 <- recipe(Weightgrams ~ Color, data = ML_linear) %>%
  step_relevel(Color, ref_level = "White") #%>% # Set reference level for White
  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors

@@

# Create a gtsummary table from model
table_model3 <- tbl_regression(ML_fit3, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs)) %>%
  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
table_model3

@@

whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.

```{r}
# Model 4: Final model for paper using all data
# Using recipes and workflows
# Complex linear model
# Specify recipe
ML_recipe4 <- recipe(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
  step_relevel(Color, ref_level = "White") %>% # Set reference level for Color
  step_relevel(ModifiedStatus, ref_level = "Multiparous") #%>% # Set reference level for ModifiedStatus
  #step_dummy(all_nominal(), -all_outcomes()) %>% # Convert all nominal variables to dummy variables
  #step_normalize(all_numeric_predictors()) # Normalize all numeric predictors

@@

# Create a gtsummary table from model
table_model4 <- tbl_regression(ML_fit4, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>% # Hide confidence intervals and p-values
  add_glance_table(include = c(r.squared, nobs)) %>% # Add R-squared and number of observations
  modify_header(statistic = "**T-Statistic**")
table_model4

### 2024-08-15 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 2
After call w/Bill

For all three models, the r-squared is very small (0.009, 0.008, and 0.026), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

@@

Now let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race). We'll start with a simple model of infant birth weight as a function of maternal skin color. Then we will run a model with all the predictors.

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)

@@

# Save regression output into table
saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))

@@

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)

# Save regression output into table
saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
```

@@

```{r}
#set seed
set.seed(123)
ML_log_split <- initial_split(ML_linear, prop = 0.75, strata = BirthweightCategory)

@@

```{r}
# Simple logistic regression model with ModifiedColor as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())

>>>>

For all three models, the r-squared is very small (0.009, 0.008, and 0.027), indicating that the model does not explain much of the variance in birth weight. This is likely due to the fact that birth weight is a complex trait influenced by many factors, including genetic, environmental, and social factors.

@@

There is potentially a (partial) multi-collinearity problem with maternal skin color and maternal nationality. So let's run a regression with just maternal nationality as the covariate.

```{r}
# Simple linear model
# Set reference level for Color
ML_linear$ModifiedNationality <- relevel(ML_linear$ModifiedNationality, ref = "Brazilian")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedNationality, data = ML_linear)

@@

# Save regression output into table
saveRDS(table_fit4, file = here::here("results", "tables", "table_fit4.rds"))
```
From our results, we can see that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, we have a problem with multicollinearity. So let's try a couple things to understand this issue. We will create a `Brazilian` (yes, no) variable and then run a regression with this variable. Then, we will drop all non-Brazilians from the study and run the original regression with the covariates.

```{r}
# Create dummy variables for new category of Brazilian or not
# Create dummy variables for status and change to English
ML_linear$Brazilian <- recode_factor(ML_linear$ModifiedNationality, "European" = "Non-Brazilian", "Middle Eastern" = "Non-Brazilian", "Latin American" = "Non-Brazilian", "Brazilian" = "Brazilian")

str(ML_linear)
```
Now, we will run a regression with only this new variable.

```{r}
# Simple linear model
# Set reference level for Brazilian
ML_linear$Brazilian <- relevel(ML_linear$Brazilian, ref = "Brazilian")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Brazilian, data = ML_linear)

# Summarize model fit with tidiers
ML_lm5_summary <- summary(ML_lm5) # Summary of fit
tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm5) # Model-level statistics

# Create a gtsummary table directly from the model
table_fit5 <- tbl_regression(ML_lm5, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit5

# Save regression output into table
saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))
```
Here, we can see that non-Brazilians have babies on average weighing 100 grams more than Brazilians (p-value 0.001). So we will now run a regression of only Brazilian women using Afro-Descent and Euro-descent categories.

```{r}
# Filter data based on desired conditions
filtered_data <- ML_linear %>%
  filter(Brazilian == "Brazilian")

# Simple linear model
# Set reference level for Brazilian
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm6 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = filtered_data)

# Summarize model fit with tidiers
ML_lm6_summary <- summary(ML_lm6) # Summary of fit
tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm6) # Model-level statistics

# Create a gtsummary table directly from the model
table_fit6 <- tbl_regression(ML_lm6, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit6

# Save regression output into table
saveRDS(table_fit6, file = here::here("results", "tables", "table_fit6.rds"))
```

Moving on from issues of collinearity, let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race). We'll start with a simple model of infant birth weight as a function of maternal skin color. Then we will run a model with all the predictors.

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)

# Summarize model fit with tidiers
ML_lm5_summary <- summary(ML_lm5) # Summary of fit
tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm5) # Model-level statistics

# Create a gtsummary table directly from the model
table_fit5 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit5

# Save regression output into table
saveRDS(table_fit5, file = here::here("results", "tables", "table_fit5.rds"))

@@

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm6 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)

# Summarize model fit with tidiers
ML_lm6_summary <- summary(ML_lm6) # Summary of fit
tidy(ML_lm6, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm6) # Model-level statistics

# Create a gtsummary table directly from the model
table_fit6 <- tbl_regression(ML_lm6, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table_fit6

# Save regression output into table
saveRDS(table_fit6, file = here::here("results", "tables", "table_fit6.rds"))
```

@@

```{r}
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW, so need to change

#set seed
set.seed(123)
ML_log_split <- initial_split(ML_linear, prop = 0.75, strata = BirthweightCategory)

@@

```{r}
# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())
